The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites

### 2023-06-11 paper/markdown/supplementary_material.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
subtitle: "Supplementary Material"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck (Innsbruck, Austria) and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study. 
The study enrollment was initiated on 10. June 2020. 
The inclusion criteria for the SIMMUN study were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck. 
The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, acute or chronic inflammatory illness and treatment with oral corticosteroids. 
The analysis inclusion criterion was the complete study variable dataset consisting of 
basic demographic and medical history variables, 
SARS-CoV-2 PCR result, titer of anti-receptor-binding domain S12/S2 protein immunoglobulin gamma (anti-RBD IgG), 
psychometric scoring of depression, anxiety and mental stress as well as blood levels of inflammatory markers and metabolites of the kynurenine and catecholamine pathway. 
A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were analyzed here. 
The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
The SIMMUN study data were gathered during a single on-site study visit at the study center at median 139 days after the SARS-CoV-2 PCR test (interquartile range: 119 - 157). 
The study visits were conducted between 17. June 2020 and 27. May 2021. 
The SIMMUN inclusion scheme is presented in __Supplementary Figure S\@ref(fig:fig-inclusion)__. 
Significant differences between participants included in the analysis and excluded due to data missingness are shown in __Supplementary Table S\@ref(tab:tab-excluded)__. 
Characteristic of the analyzed SIMMUN cohort split by the SARS-CoV-2 infection status is presented in __Table 1__.

@@

The inflammatory markers included in the analysis, plasma concentration of neopterin (NEO) measured by ELISA and neutrophil/lymphocyte ratio (NLR) were determined by a certified clinical routine laboratory at the University Hospital of Innsbruck. 
C-reactive protein (CRP) and interleukin-6 (IL6) were determined in the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants. 
Elevated CRP > 0.5 mg/dL was detected in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
elevated IL6 > 7 pg/mL was observed in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN cohort. 
Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008]. 
Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) were used as readouts of systemic indoleamine 2,3-dioxygenase (IDO, kynurenine pathway) and phenylalanine hydroxylase (PAH, catecholamine pathway) activity, respectively [@Capuron2011]. 
Plasma titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
Strongly non-normally distributed anti-RBD IgG levels were expressed as arbitrary units (AU) and stratified with the $\leq$ 1, 1 - 16.3 and > 16.3 AU cutpoints defined by the maximal titer in the uninfected participants (1 AU) and median titer in the SARS-CoV-2 subset (16.3 AU). 
Laboratory measurements at and beyond detection limits were substituted with the respective detection limit value.

Mental stress was assessed with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Clinically relevant symptoms of anxiety and depression were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
The total possible score range for each subscale is 0 to 21, with higher scores indicating more severe symptoms of anxiety/depression. 
Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed [@Bjelland2002; @Zigmond1983]. 
PSS-4, HADS anxiety and depression scales displayed good-to-excellent internal consistency as measured by the McDonald $\omega$ statistic [@McDonald1999] (__Supplementary Figure S\@ref(fig:fig-consistency)__).

In order to improve normality of some numeric SIMMUN study variables prior to linear modeling and statistical hypothesis testing with parametric tools, 
logarithm 
(`r suppl_tabs$study_vars %>% filter(Transformation == 'logarithm') %>% .[['Variable label']] %>% paste(collapse = ', ')`) or square root transformations 
(`r suppl_tabs$study_vars %>% filter(Transformation == 'square root') %>% .[['Variable label']] %>% paste(collapse = ', ')`) were applied. 
Strongly non-normally distributed HADS scores and anti-RBD IgG were stratified with the cutoffs described above. 
Transformation and stratification schemes of study variable are provided in __Supplementary Table S\@ref(tab:tab-study-vars)__.

@@

Concerning the INCOV study design, plasma metabolites and proteome were determined in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']` days), 
sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']` days) and 
recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR). 
The measurements were only partially matched by participant (i.e. multiple longitudinal measurements for an individual) hence precluding a classical repeated measurement analysis approach. 
Numbers of available INCOV cohort samples and the sampling timepoints are shown in __Supplementary Table S\@ref(tab:tab-samples)__.

@@

Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively (functions `explore()` and `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
The distribution testing revealed substantial deviations from normality for some SIMMUN study parameters. 
For this reason, logarithm and square root transformations were applied prior to modeling and statistical hypothesis testing with parametric tools as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__. 
Most of the INCOV study variables were non-normally distributed. 
For this reason, non-parametric Spearman rank test was chosen for correlation analysis and robust linear regression was used as a modeling approach.

@@

Differences in characteristic of SIMMUN participants included in the analysis and excluded due to data missingness were assessed by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Supplementary Table S\@ref(tab:tab-excluded)__). 
Differences in characteristic of SIMMUN participants or INCOV study participants divided by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Tables 1__ and __2__). 
Correlation of metabolite concentrations with age, scoring of mental stress (PSS-4) and markers of inflammation (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__). 
Pairwise correlations of cytokine and metabolite levels in the INCOV cohort were assessed by Spearman's rank test. 
Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort split by the SARS-CoV-2 infection status or presence of clinically relevant depression symptoms (HADS) were assessed by two-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).

@@

Effects of age, sex, body mass class, physical and mental disorders, BMI class, smoking and alcohol consumption, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress scoring (PSS-4) on readouts of the kynurenine (TRP, KYN, KYN/TRP) and catecholamine pathway activity (PHE, TYR, PHE/TYR) were assessed by multi-parameter linear regression with backward elimination. 
Responses and numeric explanatory variables were transformed with log or squared root functions to improve normality as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__ and normalized prior to modeling.

@@

Inference statistics for the model estimates ($\beta$, expected values and 95% confidence intervals) were extracted from the model objects with the `summary()` method. 
Significance of the estimates ($p(\beta\neq0)$) was determined by two-tailed T test. 
Results of the linear modeling in the INCOV cohort are presented in __Figure 2__ and __Supplementary Table S\@ref(tab:tab-modeling-simmun)__.

@@

Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV cohort [@Su2022]. 
We modeled normalized plasma levels of serotonin as the product of the serotonin pathway [@Lukic2022], and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA sulfate) [@Goldstein1999; @Meiser2013]. 
The explanatory variables were age, sex and BMI class, plasma levels of cytokines markers of inflammation (IL6, IL10, TNF, IFNG), plasma concentrations of metabolites of the serotonin and kynurenine pathway (TRP, KYN, QUIN), plasma concentrations of metabolites of the catecholamine pathway (PHE, TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control). 
Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.

@@

Differences in normalized blood concentrations of cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines. 
Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM `rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022]. 
Inference statistics were computed as described for multi-parameter robust linear modeling. 
The time course modeling results are presented in __Figure 4__, __Supplementary Table S\@ref(tab:tab-time-uninfected)__ - __S\@ref(tab:tab-time-acute)__.

@@

Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), metabolites of the serotonin and kynurenine pathways (serotonin, TRP, KYN and QUIN), and metabolites of the catecholamine pathway (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery). 
The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006]. 
The graphs were visualized with tools provided by the _ggnetwork_ package [@Briatte2021] as two-dimensional network plots with the node proximity determined by the $\rho$ value and distance-dependent repulsion, and edge color and width coding for the $\rho$ value and sign (__Figure 5__). 
Additionally, correlation analysis results were presented as correlograms in __Supplementary Figure \@ref(fig:fig-corr-invcov)__.

@@

flextable::flextable(suppl_tabs$incov_study_vars) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>%
  width(4, width = 6, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
   footnote(14, 2, 
           value = as_paragraph('normal: body mass index (BMI) < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of plasma concentrations of tryptophan, kynurenine, tyrosine, and kynurenine - tryptophan, and phenylalanine - tyrosine ratios in the SIMMUN cohort.'}

@@

```{r tab-univariate-correlation, tab.cap = "Correlation of age, perceived mental stress scoring, plasma neopterin levels and neutrophil/leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}

@@

```{r tab-univariate-complarison, tab.cap = "Comparison of plasma metabolite concentrations in SIMMUN study participants split by clinically relevant depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}

@@

flextable::flextable(suppl_tabs$modeling_incov) %>% 
  width(1, width = 2.5, unit = 'cm') %>% 
  width(2, width = 3.5, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4, width = 2.3, unit = 'cm') %>% 
  width(5, width = 3.9, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 1, 
           value = as_paragraph('5-HT: 5-hydroxy tryptamine/serotonin; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  footnote(1, 2, 
           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; QUIN: quinolinic acid; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine; BMI: body mass index, normal: BMI < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort. The uninfected subset served as a baseline.'}

flextable::flextable(suppl_tabs$time_uninfected) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3.7, unit = 'cm') %>% 
  width(4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.9, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 2, 
           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5, 
           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort. The acute SARS-CoV-2 infection subset served as a baseline.'}

>>>>

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
subtitle: "Supplementary Material"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

bibliography: infl_biblio.bib
csl: psychoneuroendocrinology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck (Innsbruck, Austria) and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study. 
The study enrollment was initiated on 10. June 2020. 
The inclusion criteria for the SIMMUN study were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck. 
The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, acute or chronic inflammatory illness and treatment with oral corticosteroids. 
The analysis inclusion criterion was the complete study variable dataset consisting of 
basic demographic and medical history variables, 
SARS-CoV-2 PCR result, titer of anti-receptor-binding domain S12/S2 protein immunoglobulin gamma (anti-RBD IgG), 
psychometric scoring of depression, anxiety and mental stress as well as blood levels of inflammatory markers and metabolites of the kynurenine and catecholamine pathway. 
A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were analyzed. 
The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
The SIMMUN study data were gathered during a single on-site study visit at the study center at median 139 days after the SARS-CoV-2 PCR test (interquartile range: 119 - 157). 
The study visits were conducted between 17. June 2020 and 27. May 2021. 
The SIMMUN inclusion scheme is presented in __Supplementary Figure S\@ref(fig:fig-inclusion)__. 
Significant differences between participants included in the analysis and excluded due to data missingness are shown in __Supplementary Table S\@ref(tab:tab-excluded)__. 
Characteristic of the analyzed SIMMUN cohort split by the SARS-CoV-2 infection status is presented in __Table 1__.

@@

Neutrophil/lymphocyte ratio (NLR) and serum neopterin concentration were used as markers of inflammation. 
NLR was determined by a certified clinical routine laboratory at the University Hospital of Innsbruck. 
NEO levels were measured by enzyme-linked immunosorbent assay (BRAHMS Diagnostics, Berlin, Germany). 
In addition to NLR and NEO, C-reactive protein (CRP) and interleukin-6 (IL6) were determined in the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants. 
Elevated CRP > 0.5 mg/dL was detected in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
elevated IL6 > 7 pg/mL was observed in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN participants. 
Serum concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008]. 
Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) were used as readouts of systemic indoleamine 2,3-dioxygenase (IDO, kynurenine pathway) and phenylalanine hydroxylase (PAH, catecholamine pathway) activity, respectively [@Capuron2011]. 
Serum titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
Strongly non-normally distributed anti-RBD IgG levels were expressed as arbitrary units (AU) and stratified with the $\leq$ 1, 1 - 16.3 and > 16.3 AU cutpoints defined by the maximal titer in the uninfected participants (1 AU) and median titer in the SARS-CoV-2 subset (16.3 AU). 
Laboratory measurements at and beyond detection limits were substituted with the respective detection limit value.

Mental stress was assessed with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Clinically relevant symptoms of anxiety and depression were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
The total possible score range for each subscale is 0 to 21, with higher scores indicating more severe symptoms of anxiety/depression. 
Clinically relevant symptoms of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed [@Bjelland2002; @Zigmond1983]. 
PSS-4, HADS anxiety and depression scales displayed good-to-excellent internal consistency as measured by the McDonald $\omega$ statistic [@McDonald1999] (__Supplementary Figure S\@ref(fig:fig-consistency)__).

In order to improve normality of some numeric SIMMUN study variables prior to linear modeling and statistical hypothesis testing with parametric tools, 
logarithm 
(`r suppl_tabs$study_vars %>% filter(Transformation == 'logarithm') %>% .[['Variable label']] %>% paste(collapse = ', ')`) or square root transformations 
(`r suppl_tabs$study_vars %>% filter(Transformation == 'square root') %>% .[['Variable label']] %>% paste(collapse = ', ')`) were applied. 
The strongly non-normally distributed HADS scores and anti-RBD IgG titers were stratified with the cutoffs described above. 
Transformation and stratification schemes of study variable are provided in __Supplementary Table S\@ref(tab:tab-study-vars)__.

@@

Concerning the INCOV study design, plasma metabolome and proteome analyses were performed in uninfected participants and SARS-CoV-2-infected individuals at three timepoints after diagnosis: 
acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']` days), 
sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']` days) and 
recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR). 
The measurements were only partially matched by participant (i.e. multiple longitudinal measurements for an individual) hence precluding a classical repeated measurement analysis approach. 
Numbers of available INCOV cohort samples and the sampling timepoints are shown in __Supplementary Table S\@ref(tab:tab-samples)__.

@@

Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively (functions `explore()` and `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
The distribution testing revealed substantial deviations from normality for some SIMMUN study parameters. 
Therefore, logarithm and square root transformations were applied to non-normally distributed variables prior to modeling and statistical hypothesis testing with parametric tools. 
Transformations of the SIMMUN study variables are specified in __Supplementary Table S\@ref(tab:tab-study-vars)__. 
Most of the INCOV study variables were non-normally distributed. 
For this reason, non-parametric Spearman rank test was chosen for correlation analysis and robust linear regression was used as a modeling approach.

@@

Differences in characteristic of SIMMUN participants included in the analysis and excluded due to data missingness were assessed by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Supplementary Table S\@ref(tab:tab-excluded)__). 
Differences in characteristic of SIMMUN study participants or INCOV study participants divided by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size statistic for numeric and categorical features, respectively (__Tables 1__ and __2__). 
Correlation of metabolite concentrations with age, scoring of mental stress (PSS-4) and markers of inflammation (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__). 
Pairwise correlations of cytokine and metabolite levels in the INCOV cohort were assessed by Spearman's rank test. 
Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort split by the SARS-CoV-2 infection status or presence of clinically relevant depression symptoms (HADS) were assessed by two-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).

@@

Effects of age, sex, physical and mental disorders, BMI class, smoking and alcohol consumption, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress scoring (PSS-4) on readouts of the kynurenine (TRP, KYN, KYN/TRP) and catecholamine pathway activity (PHE, TYR, PHE/TYR) were assessed by multi-parameter linear regression with backward elimination. 
Responses and numeric explanatory variables were transformed with log or squared root functions to improve normality as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__ and normalized prior to modeling.

@@

Inference statistics for the model estimates ($\beta$, expected values and 95% confidence intervals) were extracted from the model objects with the `summary()` method. 
Significance of the estimates ($p(\beta\neq0)$) was determined by two-tailed T test. 
Results of the linear modeling in the SIMMUN cohort are presented in __Figure 2__ and __Supplementary Table S\@ref(tab:tab-modeling-simmun)__.

@@

Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on kynurenine and catecholamine pathway activity were subsequently tested in the publicly available INCOV cohort [@Su2022]. 
We modeled normalized plasma levels of serotonin, and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA sulfate) [@Goldstein1999; @Meiser2013]. 
The explanatory variables were age, sex and BMI class, plasma levels of cytokines markers of inflammation (IL6, IL10, TNF, IFNG), plasma concentrations of metabolites of the kynurenine (TRP, KYN, QUIN) and catecholamine pathway (PHE, TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected individuals). 
Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.

@@

Differences in normalized blood concentrations of cytokines and metabolites between uninfected individuals, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines. 
Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM `rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022]. 
Inference statistics were computed as described for multi-parameter robust linear modeling. 
The time course modeling results are presented in __Figure 4__, __Supplementary Table S\@ref(tab:tab-time-uninfected)__ - __S\@ref(tab:tab-time-acute)__.

@@

Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), serotonin and metabolites of the kynurenine (TRP, KYN and QUIN), and catecholamine pathway (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery). 
The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006]. 
The graphs were visualized with tools provided by the _ggnetwork_ package [@Briatte2021] as two-dimensional network plots with the node proximity determined by the $\rho$ value and distance-dependent repulsion, and edge color and width coding for the $\rho$ value and sign (__Figure 5__). 
Additionally, correlation analysis results were presented as correlograms in __Supplementary Figure \@ref(fig:fig-corr-invcov)__.

@@

flextable::flextable(suppl_tabs$incov_study_vars) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>%
  width(4, width = 6, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
   footnote(14, 2, 
           value = as_paragraph('normal: body mass index (BMI) < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of serum concentrations of tryptophan, kynurenine, tyrosine, and kynurenine/tryptophan and phenylalanine/tyrosine ratios in the SIMMUN cohort.'}

@@

```{r tab-univariate-correlation, tab.cap = "Correlation of age, perceived mental stress scoring, serum neopterin levels and neutrophil/leukocyte ratio with serum levels of metabolites of the kynurenine and catecholamine pathways in the SIMMUN cohort investigated by Pearson's test."}

@@

```{r tab-univariate-complarison, tab.cap = "Comparison of serum metabolite concentrations in SIMMUN study participants split by presence of clinically relevant symptoms of depression and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}

@@

flextable::flextable(suppl_tabs$modeling_incov) %>% 
  width(1, width = 2.5, unit = 'cm') %>% 
  width(2, width = 3.5, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4, width = 2.3, unit = 'cm') %>% 
  width(5, width = 3.9, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 1, 
           value = as_paragraph('DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  footnote(1, 2, 
           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; QUIN: quinolinic acid; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine; BMI: body mass index, normal: BMI < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of plasma levels of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort. The uninfected subset served as a baseline.'}

flextable::flextable(suppl_tabs$time_uninfected) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3.7, unit = 'cm') %>% 
  width(4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.9, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 2, 
           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5, 
           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of plasma levels of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort. The acute SARS-CoV-2 infection subset served as a baseline.'}
